Unsupervised embedding of trajectories captures the latent structure of mobility

### 2021-08-07 paper/current/main.tex
update figs and merge fig

Trajectories of scientific mobility are constructed using more than three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database (see Methods).
As a scientist's career progresses, they move between organizations or pick up additional (simultaneous) affiliations forming \textit{affiliation trajectories} (Fig.~\ref{fig:methods:mobility_sentence}).
Thus, the trajectories encode both migration and co-affiliation---the holding of multiple simultaneous affiliations involving the sharing of time and capital between locations---that is typical of scientific mobility~\autocite{rodrigues2016mobility, sugimoto2017mostimpact} (see \SI).

@@

%
% Figure - Example of mobility sentences
%
\begin{figure}[ht!]
	\centering
	\includegraphics[width=0.5\textwidth]{\figdir/mobility_example_combined.pdf}
	\caption{
		\textbf{Construction of affiliation trajectories from publication records.}
		\textbf{a.}
		An author published five papers across five time periods, with only one affiliation listed in the byline of each paper.
		A unique identifier is assigned to each organization and they are assembled into an affiliation trajectory ordered by year of publication.
		\textbf{b.}
		If an author lists multiple organization affiliations within the same year, then organization IDs within that year are placed in random order in each training iteration of the \textit{word2vec} model (for more detail, see \SI).
	}
	\label{fig:methods:mobility_sentence}
\end{figure}

@@

\js{We prove that the gravity could be more powerful by using mobility-aware embedding distance between places as a distance measure. Our approach always better predicts actual mobility flows than the geographic distance and other baseline methods across three disparate datasets consistently.} In the case of flight itineraries, the embedding distance explains more than twice the expected flux between airports ($R^{2} = 0.51$, Fig.~\ref{fig:gravity_performance}a) than does geographic distance ($R^{2} = 0.22$).
Also, the embedding distance produces better predictions of actual flux between airports than does the geographic distance (Fig.~\ref{fig:gravity_performance}b).
In the case of Korean accommodation reservations, embedding distance better explains the expected flux ($R^{2} = 0.57$, Fig.~\ref{fig:gravity_performance}c) than does geographic distance ($R^{2} = 0.25$), and predictions made using the embedding distance outperform those made with geographic distance (Fig.~\ref{fig:gravity_performance}d).
This performance is consistent in the case of scientific mobility:
the embedding distance explains more than twice the expected flux ($R^{2} = 0.48$, Fig.~\ref{fig:gravity_performance}e) than does the geographic distance ($R^{2} = 0.22$), and predictions made using the embedding distance outperform those using the geographic distance (Fig.~\ref{fig:gravity_performance}f).
These patterns hold for the subsets of only domestic (within-country organization pairs, Fig.~\ref{fig:supp:predict_geo_bymodel} and Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (across-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}d).
The embedding distance also out-performs alternative diffusion-based network distance measures including the personalized-page rank scores calculated from the underlying mobility network (Fig.~\ref{fig:gravity_performance}, Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
The embedding distance derived from neural embedding also explains more of the flux and better predicts mobility flows than simpler embedding baselines, such as distances derived from a singular-value decomposition and a Laplacian Eigenmap embedding~\autocite{belkin2003laplacian} of the underlying location co-occurrence matrix, Levy's symmetric \textit{word2vec}\autocite{levy2014neural}, and even direct optimization of the gravity model (Fig.\ref{fig:gravity_performance} and Tables~\ref{supp:table:r2_table},~\ref{supp:table:rmse_exp_table},~\ref{supp:table:rmse_power_table}).
In sum, our results demonstrate that, consistently and efficiently, the embedding distance better captures patterns of actual mobility than does the geographic distance.

%
% Figure - Gravity law performance
%
\begin{figure}[h!]
	\centering
	\includegraphics[width=\textwidth]{\figdir/GravityLawFig/gravity_performance.pdf}
	\caption{
		\textbf{Embedding distance encodes functional distance and better predicts mobility in flights, accommodation reservations, and global scientific mobility compared to various baselines. }
		\textbf{a.} Embedding distance (cosine distance between organization vectors, top) better explains the expected flux of passengers between U.S. airports than does geographic distance (bottom).
		The red line is the line of the best fit.
		Black dots are mean flux across binned distances.
		99\% confidence intervals are plotted for the mean flux in each bin based on a normal distribution.
		Correlation is calculated on the data in the log-log scale ($p < 0.0001$ across all fits).
		The lightness of each hex bin indicates the frequency of organization pairs within it.
		\textbf{b.} Predictions of flux between airport pairs made using embedding distance (top) outperform those made using geographic distance (bottom).
		Box-plots show the distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps with $y = x$;
		a perfect prediction yields all points on the black line.
		``RMSE'' is the root-mean-squared error between the actual and predicted values.
		Results are consistent in the case of scientific mobility.
		For Korean accommodation reservations, embedding distance better explains the expected flux than does geographic distance (\textbf{c}), and produces better predictions (\textbf{d}).
		Similarly, in the case of global scientific mobility, embedding distance explains the expected flux between organizations (\textbf{e)} and allows for better predictions (\textbf{f}) than geographic distance.
		\js{\textbf{g.}  The correlation between the flux and baseline distances, measured with $R^{2}$, and  \textbf{h.}  the  root-mean-squared error between the actual and predicted values across various baselines.
    The asterisk denotes the top-performing method. Our approach better explains and predicts flux than any other baseline (Table~\ref{supp:table:r2_table}, Table~\ref{supp:table:rmse_exp_table}, and Table~\ref{supp:table:rmse_power_table}).}}
	\label{fig:gravity_performance}
\end{figure}

@@

%% Mobility sentences
We embed trajectories by treating them analogously to sentences and locations analogously to words.
For U.S. airport itinerary, trajectories are formed from the flight itineraries of individual passenger, in which airports correspond to unique identifiers.
In the case of Korean accommodation reservations, trajectories comprise a sequence of accommodations reserved over a customer's history.
For scientific mobility, an``affiliation trajectories'' is constructed for each mobile author, which is built by concatenating together their ordered list of unique organization identifiers, as demonstrated in Fig.~\ref{fig:methods:mobility_sentence}a.
In more complex cases, such as listing multiple affiliations on the same paper or publishing with different affiliations on multiple publications in the same year, the order is randomized within that year, as shown in Fig.~\ref{fig:methods:mobility_sentence}b.

>>>>

Trajectories of scientific mobility are constructed using more than three million name-disambiguated authors who were \textit{mobile}---having more than one affiliation---between 2008 and 2019, as evidenced by their publications indexed in the Web of Science database (see Methods).
As a scientist's career progresses, they move between organizations or pick up additional (simultaneous) affiliations forming \textit{affiliation trajectories} (Fig.~\ref{fig:gravity_performance}a).
Thus, the trajectories encode both migration and co-affiliation---the holding of multiple simultaneous affiliations involving the sharing of time and capital between locations---that is typical of scientific mobility~\autocite{rodrigues2016mobility, sugimoto2017mostimpact} (see \SI).

@@

@@

\js{We prove that the gravity could be more powerful by using mobility-aware embedding distance between places as a distance measure. Our approach always better predicts actual mobility flows than the geographic distance and other baseline methods across three disparate datasets consistently.} In the case of flight itineraries, the embedding distance explains more than twice the expected flux between airports ($R^{2} = 0.51$, Fig.~\ref{fig:gravity_performance}a) than does geographic distance ($R^{2} = 0.22$).
Also, the embedding distance produces better predictions of actual flux between airports than does the geographic distance (Fig.~\ref{fig:gravity_performance}b).
In the case of Korean accommodation reservations, embedding distance better explains the expected flux ($R^{2} = 0.57$, Fig.~\ref{fig:gravity_performance}c) than does geographic distance ($R^{2} = 0.25$), and predictions made using the embedding distance outperform those made with geographic distance (Fig.~\ref{fig:gravity_performance}d).
This performance is consistent in the case of scientific mobility:
the embedding distance explains more than twice the expected flux ($R^{2} = 0.48$, Fig.~\ref{fig:gravity_performance}e) than does the geographic distance ($R^{2} = 0.22$), and predictions made using the embedding distance outperform those using the geographic distance (Fig.~\ref{fig:gravity_performance}f).
These patterns hold for the subsets of only domestic (within-country organization pairs, Fig.~\ref{fig:supp:predict_geo_bymodel} and Fig.~\ref{fig:supp:gravity_correlation}c) and only international mobility flows (across-country organization pairs, Fig.~\ref{fig:supp:gravity_correlation}d).

%
% Figure - Example of mobility sentences
%
%
% Figure - Gravity law performance
%
\begin{figure}[!htp]
	\centering
	\includegraphics[width=0.9\textwidth]{\figdir/GravityLawFig/gravity_performance_new.pdf}
	\caption{
		\textbf{Embedding distance encodes functional distance and better predicts mobility across three distinct human trajectory datasets. }
		\textbf{a.} A unique identifier is assigned to each organization and they are assembled into an affiliation trajectory ordered by year of publication (top). If an author lists multiple organization affiliations within the same year,  we shuffle the order within that year in each training iteration (bottom, see \SI).		
		\textbf{b.} Neural embeddings outperform baselines for scientific mobility.
The correlation between the flux and baseline distances $R^{2}$ (left), and the  root-mean-squared error between the actual and predicted values (right). The asterisk indicates the top-performing method.
		\textbf{c.} Embedding distance (top) better explains the expected flux of passengers between U.S. airports than does geographic distance (bottom).
		The red line is the line of the best fit.
		%Black dots are mean flux across binned distances.
		99\% confidence intervals are plotted for the mean flux in each bin.
		Correlation is calculated on the data in the log-log scale ($p < 0.0001$ across all fits).
		%The lightness of each hex bin indicates the frequency of organization pairs within it.
		\textbf{d.} Predictions of flux between airport pairs made using embedding distance (top) outperform those made using geographic distance (bottom).
		Box-plots show the distribution of actual flux for binned values of predicted flux.
		Box color corresponds to the degree to which the distribution overlaps with $y = x$.
		%a perfect prediction yields all points on the black line.
		``RMSE'' is the root-mean-squared error between the actual and predicted values.
	Embedding distance consistently produce powerful functional distance compare to geographic distance for Korean accommodation reservations (\textbf{e,f}), and global scientific mobility (\textbf{g,h})}
	\label{fig:gravity_performance}
\end{figure}

The embedding distance also out-performs alternative diffusion-based network distance measures including the personalized-page rank scores calculated from the underlying mobility network (Fig.~\ref{fig:gravity_performance}b, Fig.~\ref{fig:supp:gravity_pprcos}, Fig.~\ref{fig:supp:gravity_pprjsd}).
The embedding distance derived from neural embedding also explains more of the flux and better predicts mobility flows than simpler embedding baselines, such as distances derived from a singular-value decomposition and a Laplacian Eigenmap embedding~\autocite{belkin2003laplacian} of the underlying location co-occurrence matrix, Levy's symmetric \textit{word2vec}\autocite{levy2014neural}, and even direct optimization of the gravity model (Fig.\ref{fig:gravity_performance}b and Tables~\ref{supp:table:r2_table},~\ref{supp:table:rmse_exp_table},~\ref{supp:table:rmse_power_table}).
In sum, our results demonstrate that, consistently and efficiently, the embedding distance better captures patterns of actual mobility than does the geographic distance.

@@

%% Mobility sentences
We embed trajectories by treating them analogously to sentences and locations analogously to words.
For U.S. airport itinerary, trajectories are formed from the flight itineraries of individual passenger, in which airports correspond to unique identifiers.
In the case of Korean accommodation reservations, trajectories comprise a sequence of accommodations reserved over a customer's history.
For scientific mobility, an``affiliation trajectories'' is constructed for each mobile author, which is built by concatenating together their ordered list of unique organization identifiers, as demonstrated in Fig.~\ref{fig:gravity_performance}a (top).
In more complex cases, such as listing multiple affiliations on the same paper or publishing with different affiliations on multiple publications in the same year, the order is randomized within that year, as shown in Fig.~\ref{fig:gravity_performance}a (bottom).
